---
title: "Minimum example-LANDFIRE data processing"
author: "Randy Swaty (actually Myles Walimaa)"
date: '2022-06-07'
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---


## Background
Due to increasingly rapid updates, need to be reproducible, desire/need to work with open-sourced tools and for the sheer fun of it we want to use R and QGIS for LANDFIRE data processing, mapping and summarization (at the least).  Recently a [Conservation Data Lab](https://conservationdatalab.org/) alumni, Myles Walimaa wrote an R script that would:

1. read in and reproject a shapefile
2. read in, mask and crop several LANDFIRE datasets (downloaded and saved as ESRI grids)
3. build attribute tables, and summary charts
4. write rasters

Subsequently, Randy Swaty added code that would:

1. make rasters readable in ArcGIS and QGIS
2. build color files for easy legend building in QGIS
3. added a few new datasets that could be read in (e.g., fuels)
4. added a couple new charts

The code Myles wrote was the backbone of assessments such as [this one](https://rswaty.github.io/deerCreekOut/deerCreekAssessment.html) completed for the Deer Creek watershed in West Virginia and for [these slides](https://rswaty.github.io/HiawathaNF2022/presentationMay2022.html#1) of Hiawatha National Forest.

While **the code is super useful** as is there is a need to review and refresh it.  Why?

* the code works with ESRI grids which are no longer delivered by LANDFIRE.  We need to test and insure it works with geotiffs
* need to update packages, e.g. "raster" to "terra" and test
* the code is verbose.  Could be made into functions
* Randy has been sloppy and has not compiled all code into one "master file".  It's time to do that.

PLUS there is an opportunity to take the code further, for example:

* could make a package
* add landscape metrics
* make maps in R so we don't have to in QGIS or ArcGIS pro.  For example make maps in LEAFLET (though these can be giant...may not be worth it, could be best to not be interactive at least as an option)
* turn this in to an R/GIS/raster training tool


## Some code examples

Here I'll share some sample code to get us started.


### Packages and reading in data


```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
# you will likely need to install many of these like:
# install.packages("tmap")

library(tidyverse)
library(sf)
library(raster)
library(foreign)
library(tmap)
library(htmlwidgets)

```



```{r read shape, message=FALSE, warning=FALSE, include=FALSE}
shp <- st_read("./DATA/deerCreek.shp") %>% 
  st_transform(crs = "+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs") %>%
  st_union() %>%
  st_sf()
```



```{r locator map, echo=FALSE, message=FALSE, warning=FALSE}
library(tmap)

# toggle tmap mode to interactive viewing
tmap_mode("view")

# create a quick interactive map
quickmap <- qtm(shp, borders = "darkgreen",
                fill = NULL)

quickmap

# save the map
# due to a bug in tmap, we need to change the working directory to OUTPUTS folder
# then switch it back when we are done saving
old.wd <- getwd()
setwd(paste0(getwd(), "/OUTPUTS"))
tmap_save(tm = quickmap, filename = "aoi_reference_map.html")
setwd(old.wd)
```

### Processing, charting and writing BpS, EVT and SCLASS data

This is older code for ESRI grids.  

## BpS


```{r bps spatial, eval=F, echo=T} 
# process BpS 
    bps <- raster("./DATA/US_200BPS/us_200bps/hdr.adf") %>%
      crop(shp) %>%
      mask(shp)
    writeRaster(bps, 
                "./OUTPUTS/bps_aoi_crop.tif", 
                overwrite = T, 
                format="GTiff", 
                options=c('TFW=YES'), 
                datatype = "INT2S")
    


    bps_x <- bps %>%
      raster::extract(shp) %>%
      unlist() %>%
      table(dnn = "ID") %>%
      as.data.frame() %>%
      arrange(desc(Freq)) %>%
      mutate_all(as.character) %>%
      mutate_all(as.integer) %>%
      left_join(levels(bps)[[1]], by = "ID") %>%
      transmute(VALUE = ID,
                COUNT = Freq,
                BPS_NAME = BPS_NAME %>%
                  as.character() %>%
                  as.factor(),
                # GROUPNAME = GROUPNAME %>%
                #   as.character() %>%
                #   as.factor(),
                GROUPVEG = GROUPVEG,
                acres = (Freq * 900 / 4046.86),
                hectare = (Freq * 900 / 10000),
                rel_percent = (Freq / sum(Freq)) * 100,
                RED = RED,
                GREEN = GREEN,
                BLUE = BLUE,
                R = R,
                G = G,
                B = B)
## write csv for fun, .dbf for use in ArcGIS
    write.csv(bps_x, "./OUTPUTS/bps_aoi_attributes.csv")
    write.dbf(bps_x, "./OUTPUTS/bps_aoi_crop.tif.vat.dbf")
    
## BpS color file for use in QGIS
    BpSColorFile <- bps_x %>%
      add_column(z = 255) %>%
      dplyr::select(VALUE,
             R,
             G,
             B,
             z,
             BPS_NAME) 
    write.table(BpSColorFile, file = "./OUTPUTS/BpSColorFile.txt", sep = ",",
                row.names = FALSE, col.names = FALSE, quote = FALSE)

```



```{r bps bar, eval=F, echo=T, fig.width=10}
# read bps attribute table .csv and summarize
bpsname <- read.csv(file = "./OUTPUTS/bps_aoi_attributes.csv") %>%
  group_by(BPS_NAME) %>%
  summarize(acres = sum(acres),
            hectare = sum(hectare),
            rel_percent = sum(rel_percent)) %>%
  arrange(desc(rel_percent)) %>%
  top_n(n = 10, wt = rel_percent)

# plot
bpsChart <- 
  ggplot(data = bpsname, aes(x = BPS_NAME, y = rel_percent)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 Biophysical Settings for Hiawatha National Forest",
    subtitle = "Represents dominant vegetation systems pre-European colonization",
    caption = "Represents ~95% of the landscape. \nData from landfire.gov.",
    x = "",
    y = "Percent of landscape") +
  scale_x_discrete(limits = rev(bpsname$BPS_NAME)) +
  coord_flip() +
  theme_bw(base_size = 14)

bpsChart

# if the plot looks good, save it
ggsave("./graphics/bpsChart.png", width = 12, height = 7)
```



## EVT


```{r evt spatial, eval=F, echo=T} 
# EVT
    evt <- raster("./DATA/US_200EVT/us_200evt/hdr.adf") %>%
      crop(shp) %>%
      mask(shp)
    writeRaster(evt, 
                "./OUTPUTS/evt_aoi_crop.tif", 
                overwrite = T, format="GTiff", 
                options=c('TFW=YES'), 
                datatype = "INT2S")
    evt_x <- evt %>%
      raster::extract(shp) %>%
      unlist() %>%
      table(dnn = "ID") %>%
      as.data.frame() %>%
      arrange(desc(Freq)) %>%
      mutate_all(as.character) %>%
      mutate_all(as.integer) %>%
      left_join(levels(evt)[[1]], by = "ID") %>%
      transmute(VALUE = ID,
                COUNT = Freq,
                EVT_NAME = EVT_NAME,
                EVT_FUEL_N = EVT_FUEL_N,
                EVT_LF = EVT_LF,
                EVT_PHYS = EVT_PHYS,
                EVT_GP_N = EVT_GP_N,
                EVT_ORDER = EVT_ORDER,
                EVT_CLASS = EVT_CLASS,
                EVT_SBCLS = EVT_SBCLS,
                acres = (Freq * 900 / 4046.86), # Freq is count of 30x30m or 900m^2 pixels, convert to acres
                hectare = (Freq * 900 / 10000),
                rel_percent = (Freq / sum(Freq)) * 100,
                RED = RED,
                GREEN = GREEN,
                BLUE = BLUE,
                R = R,
                G = G,
                B = B                )
    write.csv(evt_x, "./OUTPUTS/evt_aoi_attributes.csv")
    write.dbf(evt_x, "./OUTPUTS/evt_aoi_crop.tif.vat.dbf")
    
    ## EVT color file for use in QGIS
    EVTColorFile <- evt_x     %>%
      add_column(z = 255)     %>%
      dplyr::select(VALUE,
                    R,
                    G,
                    B,
                    z,
                    EVT_NAME) 
    
    
    write.table(EVTColorFile, file = "./OUTPUTS/EVTColorFile.txt", sep = ",",
                row.names = FALSE, col.names = FALSE, quote = FALSE)
```


```{r evt bar, eval=F, echo=T, fig.width=10}
evtname <- read.csv(file = "./OUTPUTS/evt_aoi_attributes.csv") %>%
  group_by(EVT_NAME) %>%
  summarize(acres = sum(acres),
            hectare = sum(hectare),
            rel_percent = sum(rel_percent)) %>%
  arrange(desc(rel_percent)) %>%
  top_n(n = 10, wt = rel_percent)

# plot
evtChart <-
  ggplot(data = evtname, aes(x = EVT_NAME, y = rel_percent)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 Existing Vegetation Types for Hiawatha National Forest",
    subtitle = "Represents dominant vegetation systems ca2016",
    caption = "Represents ~85% of the landscape. There are 74 EVTs total. \nData from landfire.gov.",
    x = "",
    y = "Percent of landscape") +
  scale_x_discrete(limits = rev(evtname$EVT_NAME)) +
  coord_flip() +
  theme_bw(base_size = 14)

evtChart

# save the plot
ggsave( "./graphics/evtname_barplot.png", width = 12, height = 7)
```


# SCLASS


```{r sclass spatial, eval=F, echo=T} 
# SCLASS
  

    sclass <- raster("./DATA/US_200SCLASS/us_200sclass/hdr.adf") %>%
      crop(shp) %>%
      mask(shp)
    writeRaster(sclass, 
                "./OUTPUTS/sclass_aoi_crop.tif", 
                overwrite = T, 
                format="GTiff", 
                options=c('TFW=YES'), 
                datatype = "INT2S")

    sclass_x <- sclass %>%
      raster::extract(shp) %>%
      unlist() %>%
      table(dnn = "ID") %>%
      as.data.frame() %>%
      arrange(desc(Freq)) %>%
      mutate_all(as.character) %>%
      mutate_all(as.integer) %>%
      left_join(levels(sclass)[[1]], by = "ID") %>%
      transmute(VALUE = ID,
                COUNT = Freq,
                LABEL = LABEL,
                acres = (Freq * 900 / 4046.86), # Freq is count of 30x30m or 900m^2 pixels, convert to acres
                hectare = (Freq * 900 / 10000),
                rel_percent = (Freq / sum(Freq)) * 100,
                RED = RED,
                GREEN = GREEN,
                BLUE = BLUE,
                R = R,
                G = G,
                B = B)
    
    write.csv(sclass_x, "./OUTPUTS/sclass_aoi_attributes.csv")
    write.dbf(sclass_x, "./OUTPUTS/sclass_aoi_crop.tif.vat.dbf")
    
    ## SCLASS color file for use in QGIS
    SCLASSColorFile <- sclass_x     %>%
      add_column(z = 255)     %>%
      dplyr::select(VALUE,
                    R,
                    G,
                    B,
                    z,
                    LABEL) 
    
    
    write.table(SCLASSColorFile, file = "./OUTPUTS/SCLASSColorFile.txt", sep = ",",
                row.names = FALSE, col.names = FALSE, quote = FALSE)

```

# BpS x EVC ??

```{r bpsXevc, eval=F, echo=T} 
# read bps and evt rasters
bps_r <- raster("./OUTPUTS/bps_aoi_crop.tif")
evc_r <- raster("./OUTPUTS/evc_aoi_crop.tif")

# edit the tables a little bit, mainly names
bps_att <- read.csv(file = "./OUTPUTS/bps_aoi_attributes.csv") %>%
  dplyr::select(VALUE, BPS_NAME, acres) %>%
  rename(BPS_ID = VALUE, bps_acres = acres) 

evc_att <- read.csv(file = "./OUTPUTS/evc_aoi_attributes.csv") %>%
  dplyr::select(VALUE, CLASSNAMES, acres) %>%
 rename(EVC_ID = VALUE) 

# combine rasters
bps2evc <- stack(bps_r, evc_r) %>%
  getValues() %>%
  as.data.frame() %>%
  plyr::count() %>%
  rename(BPS_ID = bps_aoi_crop, EVC_ID = evc_aoi_crop) %>%
  mutate(acres = round((freq * 900 / 4046.8564224))) %>%
  left_join(bps_att) %>%
  left_join(evc_att, by = "EVC_ID") %>%
  drop_na()


write.csv(bps2evc, "./OUTPUTS/bps2evc.csv")
```

```{r ridgeline, eval=F, echo=T} 

###  ridgeline chart
pacman::p_load(tidyverse, ggridges)

### load data
bps_evc<- read_csv("OUTPUTS/forRidge.csv")
View(bps_evc)

### format data
repeats <- bps_evc[rep(seq(nrow(bps_evc)), bps_evc$percent),]


### make chart


ggplot(repeats, aes(x = cover, y = bps)) +
  geom_density_ridges(aes(fill = bps), scale = 0.9, size = 0.5) +
  theme_ridges() + 
  theme(legend.position = "none") +
  labs(
    title = "Smoothed Tree Canopy Cover per Ecosystem",
    caption = "Data from LANDFIRE",
    x = "Canopy Cover Percentage",
    y = "" ) + 
  scale_fill_manual(values = c("#7570b3", #swamp 
                               "#d95f02", #barrens
                               "#1b9e77" )) +
  annotate(
    geom = "curve", x = 58, y = 3.7, xend = 68, yend = 3.85, 
    curvature = -.3, arrow = arrow(length = unit(2, "mm")) ) +
  annotate(geom = "text", x = 30, y = 3.7, label = "~46% of Northern Hardwoods have \nbetween 70-80% canopy cover", hjust = "left") + 
  annotate(
    geom = "curve", x = 60, y = 2.6, xend = 54, yend = 2.6, 
    curvature = .3, arrow = arrow(length = unit(2, "mm")) ) +
  annotate(geom = "text", x = 92, y = 2.5, label = "~30% of The Jack Pine-Black Spruce \nForest-Pine Barrens have \nbetween 50-60% canopy cover", hjust = "right") + 
  annotate(
    geom = "curve", x = 51, y = 1.7, xend = 58, yend = 1.85, 
    curvature = -.3, arrow = arrow(length = unit(2, "mm")) ) +
  annotate(geom = "text", x = 24, y = 1.55, label = "~45% of Alkaline Conifer-Hardwood have \nbetween 60-70% canopy cover", hjust = "left")

ggsave("./graphics/bps_evc_plot.png", width = 12, height = 6)
```